Consequences of long transients in model inference and optimal control

### 2019-10-03 manuscript/manuscript.Rmd
WIP

```{r include = FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = 1,  
                      dev = "cairo_pdf", fig.width=4.5, 
                      fig.height=3.5, message = FALSE,
                      warning = FALSE)
library(hrbrthemes)
library(ggplot2)
library(ggthemes)
library(Cairo)
library(extrafont)
extrafont::loadfonts()
```

@@

Transient dynamics have long been identified in the ecological systems

@@

>>>>

```{r include = FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = 1,  
                      dev = "cairo_pdf", fig.width=4.5, 
                      fig.height=3.5, message = FALSE,
                      warning = FALSE)
library(hrbrthemes)
library(ggplot2)
library(ggthemes)
library(Cairo)
library(extrafont)
extrafont::loadfonts()
pal <- solarized_pal()(3)
```

@@

@@

```{r results="hide"}
library(MDPtoolbox)
library(sarsop) # remotes::install_github("boettiger-lab/sarsop")
library(tidyverse) # for plotting
library(mdplearning)

damage <- 0.05
control <- 1
reward_fn <- function(x,h) - damage * x ^ 2 - control * h
discount <- 0.98

states <- seq(0,2, length=200)
actions <- states
observations <- states
sigma_g <- 0.05
sigma_m <- 0.0

r <- 0.8
K <- 1.53
q <- 2
b <- .2
eps <- states[2]/10
Tmax <- 100

may <- function(a){  
  function(x, h){ # May
    # x <- pmax(x - h, 0)   # harvest then recruit   
    x + x * r * (1 - x / K)  - a * x ^ q / (x ^ q + b ^ q) + eps - h
  }
}

possible_a <- seq(.25, .34, by = 0.005)
true_a <- 0.27   ## reality has just a transient.  use  0.277 for stronger ghost
believe_a <- 0.31 ## believe there's a strong attractor
true_i <- which.min(abs(possible_a - true_a))

```

## True model: weak ghost attractor

```{r true_a}
f <- may(true_a)
tibble(x = states[1:120],
       f = f(x,0) - x) %>%
  ggplot(aes(x, f)) + 
  geom_line(col=pal[1], lwd=2) +
  geom_hline(aes(yintercept = 0), lwd=2, lty=2) + 
  ylim(-0.06, 0.06)
```

## Incorrect belief of bistable system

```{r a29}
f <- may(0.29)
tibble(x = states[1:120],
       f = f(x,0) - x) %>%
  ggplot(aes(x, f)) + 
  geom_line(col=pal[1], lwd=2) +
  geom_hline(aes(yintercept = 0), lwd=2, lty=2) + 
  ylim(c(-0.04, 0.04))
```

## Incorrect belief of stronger bistable system

```{r a31}
f <- may(believe_a)
tibble(x = states[1:120],
       f = f(x,0) - x) %>%
  ggplot(aes(x, f)) + 
  geom_line(col=pal[1], lwd=2) +
  geom_hline(aes(yintercept = 0), lwd=2, lty=2) + 
  ylim(-0.03, 0.03)
```

```{r results="hide"}
m_true <- fisheries_matrices(states, actions, observations, reward_fn, 
                        may(true_a), sigma_g, sigma_m, noise = "lognormal")

m <- fisheries_matrices(states, actions, observations, reward_fn, 
                        may(believe_a), sigma_g, sigma_m, noise = "lognormal")
soln <- mdp_value_iteration(m$transition, m$reward, discount)
opt_soln <- mdp_value_iteration(m_true$transition, m_true$reward, discount)

```

## Simulation under *no intervention*

```{r no_intervention}
x0 <- which.min(abs(states - 0.1))
no_policy <- numeric(length(states)) + 1

df <- mdp_planning(m_true$transition, m_true$reward, discount, model_prior = c(1), 
                   policy = no_policy, x0 = x0, Tmax = 100)
```

```{r}
df %>% mutate(state = states[state], action = 5*actions[action]) %>% 
  select(state, action, time) %>% 
  gather(series, state, -time) %>%
  ggplot(aes(time, state, col=series)) +
  geom_line(lwd=2) + 
  ylim(0,1.3)
```

## Mistaken ghost for alternate stable state

```{r mistaken_ghost}
x0 <- which.min(abs(states - 0.1))
no_policy <- numeric(length(states)) + 1

df <- mdp_planning(m_true$transition, m_true$reward, discount, model_prior = c(1), 
                   policy = soln$policy, x0 = x0, Tmax = 100)

df %>%
  mutate(state = states[state], action = 5*actions[action]) %>% 
  select(state, action, time) %>% 
  gather(series, state, -time) %>%
  ggplot(aes(time, state, col=series)) +
  geom_line(lwd=2) + 
  ylim(0,1.3)
```

## Prior beliefs

```{r priors}
prior <- dnorm(possible_a, 0.295, 0.005)
prior <- prior / sum(prior)
data.frame(a = possible_a, probability = prior) %>%
  ggplot(aes(a,prior)) + 
  geom_bar(stat="identity", fill=pal[1], alpha=0.6) +
  geom_vline(aes(xintercept = true_a), lwd=3, lty=2, col=pal[1]) + 
  ylim(0, 0.6)
```

## Learning late

```{r learning}
sim <- readRDS("../notebook/sim-ghost-learning.Rds")
states <- seq(0,2, length=140)
actions <- states
Tmax <- 100

 sim$posterior %>% 
  data.frame(time = 1:Tmax) %>%
  filter(time %in% seq(1,Tmax, by = 10), time < 50) %>%
  gather(param, probability, -time, factor_key =TRUE) %>% 
  mutate(param = as.numeric(param)) %>% 
  ggplot(aes(x = param, y = probability, group = time, alpha = time)) + 
  geom_bar(stat="identity", position = "identity", show.legend = FALSE, fill=pal[1]) 
```

## Outcome

```{r}
 sim$df %>% 
   mutate(state = states[state], action = 5*actions[action]) %>% 
  select(-value, -obs) %>% 
  gather(series, state, -time) %>% 
  ggplot(aes(time, state, color = series)) +
  geom_line(lwd=2) +
  ylim(0,1.3) +
  ylab("state / action")
```

## More agnostic prior beliefs

```{r priors_wide}
prior <- dnorm(possible_a, 0.3, 0.05)
prior <- prior / sum(prior)
data.frame(a = possible_a, probability = prior) %>%
  ggplot(aes(a,prior)) + geom_bar(stat="identity", fill=pal[1], alpha=0.5) +
  geom_vline(aes(xintercept = true_a), lwd=3, lty=2, col=pal[2]) +
  ylim(0,.3)
```

```{r less-learning}
sim2 <- readRDS("../notebook/sim-ghost-learning-2.Rds")
states <- seq(0,2, length=200)
actions <- states
Tmax <- 100

 sim2$posterior %>% 
  data.frame(time = 1:Tmax) %>%
  filter(time %in% seq(1,Tmax, by = 10), time < 50) %>%
  gather(param, probability, -time, factor_key =TRUE) %>% 
  mutate(param = as.numeric(param)) %>% 
  ggplot(aes(x = param, y = probability, group = time, alpha = time)) + 
  geom_bar(stat="identity", position = "identity", show.legend = FALSE, fill=pal[1]) + 
   ylim(0,0.3)
```

## Less learning, more caution

```{r}
 sim2$df %>% 
   mutate(state = states[state], action = 5*actions[action]) %>% 
  select(-value, -obs) %>% 
  gather(series, state, -time) %>% 
  ggplot(aes(time, state, color = series)) +
  geom_line(lwd=2) + 
  ylim(0,1.3) +
  ylab("state / action")
```

```{r}
```
